First results of developing a ruled based Swedish parser

### 2011-10-09 doc/paper/Paper.tex
paper

\documentclass[submission,copyright,creativecommons]{eptcs} %rätt lincens?
\providecommand{\event}{LREC2012}
%\usepackage{breakurl}        
\usepackage{graphicx}

\title{Den stora grammatiken} % ska vi vara anonyma?
\author{Malin Ahlberg
\institute{Gothenburg University, Sweden}
\email{ahlberg.malin@gmail.com}
}
\def\titlerunning{First results}
\def\authorrunning{M. Ahlberg}
\begin{document}
\maketitle

\begin{abstract}
om parsning of free language (take from old report)
but for linguistics, make a wide coverage grammar, robust parser
By using existing technologies
and tools:use three resources, gf, talbanken, saldo
A parser like this would be of great use for many natural
language processing applications, such as translation and information retrieval
as well as making semantic representations.
The goal is to be able to parse open domain language and the
parser will be evaluated on an extensive Swedish
treebank, Talbanken.

This report describes the start of implementing a robust parser for Swedish,
using the grammar formalism Grammatical Framework\cite{gf} (GF).

@@

\section{Introduction}

Making computers able of handling human language is a 
hard problem.
The meaning of a sentence depends not only of which words it consists of, but
also on their syntactic use, how they interact and relate to each other.
For a computer to make sense of natural language, it needs to analyse this 
syntactic structure; it needs a good grammar and parser.
Moreover, the parser needs a preprocessor, named entity recognizer and more
tequnices to make it robust.

We are working on an extended implementation of a grammar from Swedish, allowing
complicated constructions such as focusing different parts of the sentences,
using complex relative clauses etc and constructions (byt ord) specific to Swedish.

The libraries of GF provide a basic grammar for
Swedish, covering the fundamental features of the language, such as morphology and
commonly used syntax. This is suitable 
for building domain specific applications. In those cases the user is not
allowed to freely compose sentences, but has to stay within the bounds of a 
controlled language. Both the vocabulary and grammatical structures are fixed,
meaning that there only is a limit number of ways to write a sentence.
Parsing open domain natural language is a much bigger task, since it involves handling both
standard and non-standard grammatical constructions.

The future work will include extending and enhancing the existing Swedish GF grammar,
importing lexicon and develop techniques for handling unknown words and grammatical
constructions, proper names, idioms, ellipses etc, in order to make the parser robust.

By using this, we hope to eventually implement a grammar able to handle free % synonyyym!
Swedish. It would be of great use for many natural
language processing applications, such as translation and information retrieval
as well as making semantic representations.

\subsection{Grammatical Framework}

@@

% about why we need more, extra module, separate swedish from scandinavian
GF\cite{gf} (Grammatical Framework) is a grammar formalism based on functional programming,
and suitable for multilingual grammar applications. The key idea is to have
one abstract grammar, modelling the structure of the domain. 
In addition, concrete grammars are implementing the abstract in different languages.
Since all languages supported in an application shares a common abstract syntax,
multilingual is easy to support.

In this way the grammar becomes independent, it may be reused in another project
and the implementation may be changed without modifying the application utilizing the grammar.

GF provides resource grammars\cite{resource} implementing information about the morphological and
syntactical rules for 18 different languages. By utilizing these, 
GF is an advantageous tool for writing special-purpose grammars.

The libraries of GF, the resource libraries\cite{resource}, provides
a common abstract and implementations of this in 18 languages.
Since translation always should be possible, the abstract may only contain constructions that
are common to all implemented languages.

  But this also means that all ---- added to the resources
  must be translatable to and expressable in all other language.
 therefor, the resources are very general, non language specific.

Language specific constructs may be
given in the module \verb|Extra.gf|. This may be stylistic changes, idioms, informal
expressions etc.
As an example, the Swedish \verb|Extra| module contains a function for expressing sentences
where the negation is put in focus: \emph{``Inte var jag glad"} (\emph{``Not was I happy"}).

The work to enhance and expand the Swedish GF grammar has already been started
% parsing free language, more work is required and it has been investigated what
%Further, other tools that are needed have been implemented and examined.
%will be needed to make it sufficient and a start on developing tools has beenmade.

\section{Talbanken}
%Wherefrom, how big, who, when, version, used for maltparser
Talbanken\cite{talbanken} is
a Swedish treebank which was put together in the 1970s at Lund University,
consisting of 6316 sentences. Each word is tagged with 
its part of speech, syntactic function and its position in a head dependent
analysis.
In 2005 Talbanken was mordenized \cite{} and enriched with annotation for a
full phrase structure analysis. Talbanken05 was used when training the 
data-driven parser Maltparser \cite{}, which var bra?

\section{Saldo}

SALDO\cite{saldo} is an open source lexicon resource
based on Svenskt Associations Lexikon (SAL). It is
developed at Sprakbanken\cite{} at Gothenburg University
and intended to be used in language technology
research. 
% how it is tagged, other usages, benefits,
% how easy it is to map it to gf
From SALDO, a large GF lexicon, \\
\verb|DictSwe.gf| has earlier been extracted \cite{},
containing 50 000 entries. Using the same updated tecniques,
we would like to reimport the lexicon in order to enlarge,
enhance it. The importing-method should be fast and reliable
enough to be able to always have an fresh version of the dictonary
in GF.
We are using parts of Talbanken for testing and development,
and the parsetree generated by our grammar will eventually be 
compared to those from the treebank.

\section{The lexicon tool}
A tool for automatically acquisition has been created. Making use of the
tags in Talbanken and the paradigms in the Swedish resource grammar, 
it interactivelly generates GF lexicons. The user
verifies tho correctness of a guess from the program, and can either
allow the word to be added to the lexicon, remove it or demand the program
to make another guess.
Altough using very simple tecniques, the test of the tool on verbs 
has a correctness rate of about
70-75\%, and can easily ? be improved.

use together with Extract/FM,

\section{Mapping of trees}
The information from the tags in Talbanken can be used for many xx.
The tools for lexical extraction can be enhanced if we get more information
about which form the word is currently used in. 
We are currently working on a automatical mapping from Talbanken trees to trees in GF format.
This would later enable us to extract possibilities for how often different functions
are used, a feature (!) that would improve the parsing considerably.
The evalution of the parser could also be accomplished by comparing the trees
from the parser and from the mapping-process.
brastartord there are differences in the notation of the sättet träden är uppbyggda på \\
 exempel?? \\

explation of the trees
why we need the mapping : evaluation, adding words, finding missing constructions
difficulties

\section{New things and hard things in the grammar}
Due to the ressemblemences between the Scandinavian languages, they share about 80\% (?)
of their resource code. This is good because blabla.
However, for a bigger Swedish grammar, the implementation of the languages need to be more separated,
in order to avoid ad-hoc solutions and constructions that are standard in
Norweigan and Danish but very rare in Swedish. \\ %skriv om här änna
A number of grammatical constructions have been added the the Swedish grammar.
Those include using the reflexive pronoun \emph{sitt}:\\
\textbf{Han såg sitt hus} \emph{He saw his (own) house} \\
as opposed to \\
\textbf{Han såg hans hus} \emph{He saw his (an other person's) house} \\
% to simple for linguists?
The possibility to put a part of a sentence in focus has also been added such as:\\
\textbf{Glad var han inte} \emph{Happy was he not}
Some ellipses are added, while there is work going on to add trickier ones (eller inte 
ellipser just nu dårå, men vi skriver något fint här med, ja-jaaaa!).

\section{Future Work}
robust parser, ner, preprocessing, adding new words,
probabilties
ellipses, chunk parsing

\section{Conclusion}
hard but nice
why better than statistical

\nocite{*}
\bibliographystyle{eptcs}
\bibliography{ourBib}
\end{document}

>>>>

\documentclass[submission]{eptcs} %rätt lincens?
\providecommand{\event}{LREC2012}
\usepackage[utf8]{inputenc}
%\usepackage{breakurl}        
\usepackage{graphicx}

\title{First results of developing a ruled based Swedish parser} %A Wide-Coverage Swedish Grammar} % ska vi vara anonyma?
\author{Malin Ahlberg
% vad ska stå här eg?
\institute{Center of Language Technology, Gothenburg University, Sweden}
%\email{ahlberg.malin@gmail.com}
}
\def\titlerunning{First results of developing a ruled based Swedish parser}
\def\authorrunning{M. Ahlberg}
\begin{document}
\maketitle

\begin{abstract}
This paper describes an on-going work of a rule-based, open source
parser for Swedish, based on a wide-coverage grammar implemented in
Grammatical Framework (GF).
In addition to GF, we use two other main resources; the Swedish treebank Talbanken
and the electronic lexicon SALDO.
The grammar will be highly reusable as GF provides libraries for compiling
grammars to a number of programming languages.
% allow easy interaction with other programs and grammars
%

%>> something very commercial, interesting and outstanding here please!

%R: this phrase "We aim to implement a parser for open domain language
%and this paper describes our work and results so far as well as the
%directions of future work." doesn't belong in the abstract because it
%just describes the normal structure of any paper, without saying why
%your work are interesting/outstanding, worth-reading-further. Make it
%sound more commercial, if you want to have it there.

@@

\section{Introduction}
%R: the first sentence - in "deep robust parser" - why "deep" - grammar
%formalism, why "robust" - ??
% afterwards, you don't actually have named-entity recognition, and
%further more, there's no such thing as "exhaustive lexicon" - you can
%always switch to another domain and find new words. Further on, the
%work has nothing to do with statistitics so far, so you can't mention
%as opening line.  The same goes for parallel chunk parsing.

%R: the first sentence doesn't really make sense because you first
%refer to a parser and only then to GF and a grammar, but without
%giving an explanation of how making the grammar would give you a
%parser for free.

%R: -- the second paragraph - OK but please extend.
We would like to implement a wide coverage grammar and parser for Swedish.
We thereby investigate how the grammar formalism Grammatical Framework,
further described in section \ref{sec:gf}, can be used for open domain parsing.
By starting from an already existing Swedish grammar written in GF, we get 
an fundamental description of the language. The framework also provides
tools such as parsing, generation and
a well tested interpretation of the parse trees. There are also tools for using
the grammar in a number of programming languages. 
%>> , tools for visualizing ...

From the GF grammar we aim to implement a robust parser. 
The parser will be evaluated both by experts and by comparing the results of
parsing sentences in an often used Swedish treebank, Talbanken.

%We aim to implement % The goal is to be able to 
%a parser for open domain language and 
%This paper describes our work and results 
%so far as well as the directions of future work.

\section{Background}
\subsection{Grammatical Framework}
\label{sec:gf}

%R: the GF part - "especially suitable" - it's actually meant for the purpose
%:)

@@

Grammatical Framework\cite{gf} (GF) is a grammar formalism based on functional
programming, designed for multilingual grammar applications.

The key idea
is to divide the grammar into abstract and concrete.
The abstract grammar gives a logical representation of the semantics of the
domain, modeled as abstract trees. Complications occurring in
natural languages, such as agreement, case and word order, are abstracted away. 
%R: what complications do you abstract away ?
%categories and trees are not interesting as such, but their usage
%is - why do you need them in the abstract syntax actually ?
That is, the abstract grammar contains no actual linguistic information, but
information about what categories are used and how to combine them into trees.

To add a new languages means to give a concrete grammar. This implements
the abstract grammar and describes how to translate abstract
trees to strings of the language. The framework enables us to parse strings into 
abstract trees as well as linearize trees into strings.
Since all languages supported in an application shares a common abstract syntax,
we get multilinguality for free.

% obra mening
The grammar acts as an independent module and may be used in different
projects.
%R: more about the GF library - if you have space
Reusability is further supported by the GF resource grammars\cite{resource},
a library provided with the GF package, 
where information about the morphological and
syntactical rules for more than 20 different languages are implemented.
As a result, an end user of the framework does not need to describe how to 
form standard sentences, phrases or decline words, but can focus on the
particularities for the given domain and choose what kind of 
utterances that should be allowed.

%so far bara used för vissa saker, 
% about why we need more, extra module, separate swedish from scandinavian
% get inspired by Shafqat.
GF has so far been used in a number of projects, MOLTO\cite{molto} and
WebAlt\cite{webalt} to mention a few. 
All those are special domain applications, dealing with controlled natural
language.
This project experiments on other usages of a GF grammar.

% parsing free language, more work is required and it has been investigated what
%Further, other tools that are needed have been implemented and examined.
%will be needed to make it sufficient and a start on developing tools has beenmade.

\subsection{Talbanken}
%Wherefrom, how big, who, when, version, used for maltparser
Talbanken\cite{talbanken} is
a Swedish treebank put together in the 1970s at Lund University.
In 2005 it was modernized by Nivre, Nilsson and Hall\cite{talbanken05} and
enriched with annotation for a
full phrase structure analysis. 
Talbanken contains both written and spoken Swedish. For this project only
the prose material, consisting of 6316 sentences will be used.
This part was also used when training the data-driven parser Maltparser \cite{malt}. \\
We are using parts of Talbanken for testing, development and evaluation.
%and the parsetree generated by our grammar will eventually be 
%compared to those from the treebank.

\subsection{Saldo}
%R: Saldo part - why would you like to reimport ? "enlarge and enhance"
%- how do you enhance it ?
SALDO\cite{saldo} is an open source lexicon resource
based on Svenskt Associationslexikon (SAL). It is
developed at Språkbanken at Gothenburg University
and intended for usage in language technology
research. 
% how it is tagged, other usages, benefits,
% how easy it is to map it to gf
From SALDO, a large GF lexicon
has earlier been extracted,
containing 50 000 entries.
By updating this technique, we would like
to reimport SALDO, which has since then been developed.
%and we would like
%to reimport the it using the same
%updated techniques. %in order to enlarge our lexicon,
%we  in order to enlarge and
%GF version. 
The importing method should be fast and reliable
enough to be allow us to always have a fresh version of the dictionary
in GF.

\section{Work in progress}
\subsection{A tool for lexical acquisition}
%R: lexicon extraction part - more about everything, comparison with
%Extract and Functional Morphology, evaluation of the lexicon builder -
%"supervised lexicon extraction"(common term for this - google it if
%you want to find related work)

To enlarge the lexicon, 
a tool for automatically acquisition has been created. It has 
been tested on verbs with good results. It makes use of
the \emph{smart paradigm} given in the Swedish resource grammar.
The smart paradigm accepts verbs in present tense indicative form and decides
 which paradigm it most likely belong to. If needed, it also accepts more
verb forms showing the correct inflection.
%which Making use of the

By combining this method with the information from the tags in Talbanken,
%and the paradigms in the Swedish resource grammar, 
the tool interactively generates GF lexicons. 
Given a list of words, it iteratively
tries to figure out how to conjugate each of them. If several forms of a word is 
given, the program will try to identify the one that carries the most linguistic
information, put this in a form recognized by the smart paradigm and ask GF to output
a table with the resulting inflection table. 
If this table contains all other forms from the input list,
the program will ask the  user to
validate the claimed conjunction. The user may now either
allow the word to be added to the lexicon, remove it or demand the program
to make another guess.
Although using simple techniques, the tool 
manage to correctly guess 70-75\% of the given lemmas when tested on
verbs. \\
For further extractions, there are also earlier developed tools such as
Extract\cite{extract} for supervised lexicon extraction and
FM\cite{fm} for programming lexical resources. Extract uses a similar 
method but does not have access to the GF smart paradigm.

%% Write more here!!
\subsection{Mapping of trees}
%R: mapping - very nice - more about it
The information from the tags in Talbanken can be used for many purposes.
We are currently working on an automatic transformation of Talbanken trees 
to trees in GF format. The translation makes use of the POS tags as well as
the syntactic information and the mapping has so far turned out to be unambiguous.

The information given by this mapping may be used for the lexical extraction
tools. Those can be enhanced if they are given more data about which form a word is
currently used in. 
The translated trees will also enable us to extract possibilities for how often
different functions are used, a feature that would enable disambiguation.
Furthermore, the translation makes it easy to identify grammatical constructions
missing from the GF grammar and shows how the GF analysis differs from the one made
in Talbanken.

%>> by looking on how often a gf function is used
%<< that would improve the parsing considerably.
Another important use of the mapping is evaluation. By comparing the trees
from the parser and from the transformer, we get an interesting way of
accomplishing an evaluation.
%An evaluation of the parser
%could also be accomplished by  \\
%<< brastartord there are differences in the notation of the sättet träden är uppbyggda på \\
%>>maybe something about how it's going and why it's hard?
%>> exempel?? \\

%explation of the trees
%why we need the mapping : evaluation, adding words, finding missing constructions
%difficulties

\subsection{Development of the grammar}
%R: New things and hard things in the grammar" - this goes hand in
%hand with more explanations about resource library and functors for
%families, also emphasize the words that you refer to in the example -
%bold font or smth equally noticeable. Explain more why these
%constructions weren't there before.

For multilingual GF applications, translation should always be possible, and
therefore the abstract may only contain constructions that
are common to all implemented languages.
%>> resources for controlled language. does not aim for covering whole language
%>> googla efter bra citat
This means that everything added to the resources
must be expressable in all other languages.
Therefore, the resources have to be general and non language specific.
Language specific constructs such as stylistic changes, idioms, informal
expressions etc may be given in a special module.
This module, together with the Swedish resource grammar has been the starting
point of this project.

Since many of the languages in the GF library reassemble each other grammatically,
they can share much of their grammar implementations. This is usually done by using a
\verb|Functor|, which lets a number of languages share parts of their implementation. 
In addition to simply avoiding code duplication, this technique aids the code maintenance.
For Swedish, about 85\% of the resource code is shared with the other Scandinavian
languages.
%due to the resemblances between the languages, they share
%>> using a functor for sharing code, for when having the same structure but
%>> but only different words. 
For a bigger and still well-defined Swedish grammar however, the implementation of
the languages needs to be more separated. % in order to avoid ad-hock solutions.
%<<in order to avoid ad-hoc solutions and 
Some constructions that are standard in
Norwegian and Danish may be very rare in Swedish and while extending the
grammar there may occur situations where a deeper restructuring of the
implementation is needed. \\
A number of grammatical constructions have been added the the Swedish grammar.
%>> this is language specific, therefore not in common abstract
Those include constructs that are generally not present in other languages,
such as the reflexive pronoun \emph{sitt}: \\
\emph{Han såg \textbf{sitt} hus} $\;$ (\emph{He saw his (own) house}) $\;$
as opposed to \\
\emph{Han såg \textbf{hans} hus} $\;$ (\emph{He saw his (an other person's) house}) \\
%>> an stylistic change that has been added, not needed for expressable new things (typ) 
%>> not critical, does not have to be in resource grammar
%>> also not expressable in all languages
The possibility to put a part of a sentence in focus has also been added:\\
\emph{Glad var han inte.} $\;$ (\emph{Happy was he not}). \\
This sort of rephrasing which has little effect on the logical representation
is commonly not given by the resource grammar.

In order to keep the grammar as clean and to keep it from being too allowing,
much care is taken to develop this as neatly as possible. 
%<<Some ellipses are added, while there is work going on to add trickier ones (eller inte 
%<<ellipser just nu dårå, men vi skriver något fint här med, ja-jaaaa!).

%% Rephrase
\section{Evaluation}
%R: you need especially an evaluation part where you can also argue
%that the method is better than statistics.
Compared to a statistical parser which operates like a black box,
a rule-based one is not only theoretically interesting,
but could also give a more explanatory output since the rules
are given their names by a human and hence acts as informative labels.
Chosen parts of the parser or grammar may also be used in other applications,
which may be dealing with controlled natural languages. Automatic language
generation from the grammar is also provided by GF.

We intend to evaluate the parser both automatically - by comparing the output
the translated trees from Talbanken - and manually by an expert in Swedish grammar.

%% more explaination of probabilities
\section{Future Work}

After having developed the grammar and lexicon, we plan to 
make the parser robust by using techniques such as chunk parsing, named
entity recognition, methods for handling unknown words and
grammatical constructions, idioms, ellipses etc.

It is already possible to add probabilities for GF functions,
which serves as a ranking when several parse trees are possible.
By adding dependency probabilities, we would like to improve the
disambiguation.

% ha med det som inte nämnts innan
% ha en jämförelse med stat.parse?
% är den deep?
\section{Conclusion}
We intend to implement a robust deep parser for Swedish, by first developing
a large scale grammar and then equip this with named entity recognition, a
voluminous lexicon, statistical information for disambiguation,
a method of parallel chunk parsing etc.
So far we have been working on the grammar, the lexical resources and
a translation of the treebank Talbanken to GF.
The usage of Grammatical Framework gives us the advent age to start from
a well-defined system of describing grammar, as well as tools for
using the grammar in combination with programming languages like Haskell
and Java.

\section{Acknowledgments}
Special thanks to Aarne Ranta, Ramona Enache, Elisabet Engdahl, Krasimir
Angelov and Lars Borin for their help and support.

\nocite{*}
\bibliographystyle{eptcs}
\bibliography{ourbib}
\end{document}
